Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-08-24 manuscript/genome_research/manuscript_genome_research.tex
added genome research submission

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\usepackage{lmodern} % Chloe: better font rendering
\usepackage[shortcuts]{extdash} 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3,4,\spadesuit}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, \\
GENESIS Study collaborators$^{1,2,3}$, Dominique Stoppa-Lyonnet$^{5,6,7}$, \\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{\small{
  \raggedright{
  $^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, 103-0027, Japan; \\
  $^{5}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{6}$INSERM, U830, F-75005 Paris, France;\\
  $^{7}$Université Paris Descartes.\\
  }
  \raggedright{$^{\spadesuit}$Corresponding author: \href{hector.climente@riken.jp}{hector.climente@riken.jp}\\}
}
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. based on logistic regression). Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

Network-based biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

>>>>

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[round]{natbib}
\usepackage{lmodern} % Chloe: better font rendering
\usepackage[shortcuts]{extdash} 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, \\
GENESIS Study collaborators$^{1,2,3}$, Dominique Stoppa-Lyonnet$^{4,5,6}$, \\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{\small{
  \raggedright{
  $^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes.\\
}}}

\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \citep{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. based on logistic regression). Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \citep{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \citep{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \citep{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as is the case in GWAS \citep{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \citep{barton_infinitesimal_2017}. The recently proposed omnigenic model \citep{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \citep{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \citep{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \citep{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \citep{huang_systematic_2018}.

Network-based biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \citep{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \citep{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \citep{jia_dmgwas:_2011} and heinz \citep{dittrich_identifying_2008}, focus on larger subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \citep{leiserson_pan-cancer_2015}, SConES \citep{azencott_efficient_2013}, and SigMod \citep{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We worked on the GENESIS dataset \citep{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.7\linewidth]{../figures/sfigure_2.png}
  \caption{Association in GENESIS. The red line represents the Bonferroni threshold. \textbf{(A)} SNP association, measured from the outcome of a 1 d.f. $\chi^2$ allelic test. Significant SNPs that are within a coding gene, or within 50 kilobases of its boundaries, are annotated. The Bonferroni threshold is 2.54 \texttimes{} 10\textsuperscript{-7}. \textbf{(B)} Gene association, measured by P\=/value of VEGAS2 \citep{mishra_vegas2:_2015} using the 10\% of SNPs with the lowest P\=/values. The Bonferroni threshold is 1.53 \texttimes{} 10\textsuperscript{-6}. \textbf{(C)} SNP association as in panel (A). The SNPs in black are selected by a L1-penalized logistic regression (Section~\ref{methods:classifier}, $\lambda = 0.03$).}
  \label{sfig:snp_gene_manhattan}
\end{figure}

We conducted association analyses in the GENESIS dataset at both SNP and gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (Figure~\ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \citep{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \citep{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \citep{rinella_genetic_2013}. In our studied population, only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (Figure~\ref{sfig:snp_gene_manhattan}B).

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \citep{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and new ones (Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Figure~\ref{fig:benchmark}B), while providing more interpretable solutions.

\begin{figure}[htbp]
  \centering
  \includegraphics[height=.75\textheight]{../figures/figure_1.png}
  \caption{Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. As LEAN did not produce any significant gene (BH adjusted P\=/value~<~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)} Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(C)} Precision and recall of the evaluated methods with respect to Bonferroni-significant SNPs/genes in BCAC. For reference, we added a gray line with a slope of 1. \textbf{(D)} Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (BH adjusted P\=/value < 0.05), we obtained six solutions (Figure~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Figure~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Figure~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes: protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, dmGWAS solution is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \citep{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

Despite their differences, there are additional common themes. All obtained solution subnetworks have lower association P\=/values than the whole PPIN (median P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Figure~\ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solution subnetworks presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value < 0.03). Second, the genes in four solution subnetworks displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value < 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \citep{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (Table~\ref{tab:snp_solutions}), suggesting that causal SNPs are also more central than non associated SNPs.

\subsection{A case study: the consensus network}
\label{results:consensus}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{../figures/figure_3.pdf}
  \caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in Figure \ref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.675\linewidth]{../figures/sfigure_4.pdf}
  \caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). \textbf{(A)} Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset. The latter ones are also colored in pink. This panel is identical to Figure~\ref{fig:consensus}. \textbf{(B)} Same network, but every gene name is indicated.}
  \label{sfig:consensus-names}
  \end{figure}

Despite the heterogeneity of the solutions, their shared properties suggest that each method captures different aspects of cancer susceptibility. Indeed, only 20 genes are common to more than two solutions (Figure~\ref{sfig:consensus_stats}A), but encouragingly, the more methods selected a gene, the higher its association score to the phenotype (Figure~\ref{sfig:consensus_stats}B). To leverage on their strengths and compensate their respective weaknesses, we built a consensus subnetwork that captures the mechanisms most shared among the solution subnetworks (Section~\ref{methods:methods}). This subnetwork (Figure~\ref{fig:consensus}) contains 93 genes and exhibits the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.65\textwidth]{../figures/sfigure_5.pdf}
  \caption{Genes on the consensus network. Breast cancer susceptibility genes are colored in pink; the rest are colored in grey. \textbf{(A)} Number of methods selecting every gene in the subnetwork. \textbf{(B)} VEGAS2 P\=/values of association of the genes, with regards to the number of methods that selected them. \textbf{(C)} Comparison of betweenness centrality of the genes in the consensus network and the other genes in the PPIN and not in the consensus network. To improve visualization, we removed outliers. \textbf{(D)} Relationship between the log\textsubscript{10} of the betweenness centrality and the -log\textsubscript{10} of the VEGAS2 P\=/value of the genes in the consensus network. The blue line represents a fitted generalized linear model.}
  \label{sfig:consensus_stats}
  \end{figure}

A pathway enrichment analysis of the genes in the consensus network also shows similar pathways as the individual solutions. We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \citep{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \citep{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to attenuation phase of heat shock response, it involves three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 map to all of them, and 4 to two, making hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

Topologically the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The opposite would be the case of highly central genes in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, Figure~\ref{sfig:consensus_stats}D). This suggests that they were selected because they were on the shortest path between two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \citep{boyle_expanded_2017}. For instance, the most central node in the consensus network is COPS5, a component of the COP9 signalosome which regulates multiple signalling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \citep{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus subnetwork have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

We compared the results of different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \citep{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a family history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network metods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC, in an equivalent way to the one described in Section~\ref{methods:node_score}. The solutions provided by the different network approaches overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based network methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (Figure~\ref{fig:solution_overview}C). Interestingly, while SConES GI at the SNP-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

\begin{figure}[htbp]
\centering
\includegraphics[width=.72\linewidth]{../figures/figure_4.pdf}
\caption{Comparison of network-based GWAS methods on GENESIS. Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:comparison}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)} Number of SNPs selected by each method and number of SNPs in the active set found by the classifier. Points are the average over the 5 runs; lines represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison. For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(B)} Sensitivity and specificity on test set of the L1-penalized logistic regression trained on the features selected by each of the methods. In addition, the performance of the classifier trained on all SNPs is displayed. Points are the average over the 5 runs; lines represent the standard error of the mean. \textbf{(C)} Pairwise Pearson correlations of the solutions used by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)} Runtime of the evaluated methods, by type of network used (gene or SNP). For gene network-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
\label{fig:benchmark}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=.8\linewidth]{../figures/figure_2.pdf}
\caption{Drawbacks encountered when using network guided methods. \textbf{(A)}~DmGWAS solution subnetwork. Genes with a P\=/value < 0.1 are highlighted in red. \textbf{(B)}~Centrality degree and -log\textsubscript{10} of the VEGAS2 P\=/value in BCAC for each of the nodes in the PPIN. We highlighted the genes selected by each method, and the ones selected by more than one (``Consensus''). We labeled the three most central genes that were picked by any method. \textbf{(C)}~Overlap between the solutions of SConES GS, GM or GI in the different genomic regions. SNPs that were not selected in the studied network, but were selected in another one, are displayed in background color. }
\label{fig:issues}
\end{figure}

In this dataset, heinz's solution is very conservative, providing a small solution with the lowest median P\=/value for the subnetwork (Table~\ref{tab:gene_solutions}). Due to this parsimonious and highly associated solution, it was the best method to stably select a set of biomarkers (Figure~\ref{fig:benchmark}C). Its conservativeness stems from its preprocessing step, which models the gene P\=/values as a mixture model of a beta distribution and a uniform distribution, controlled by an FDR parameter. Due to the limited signal at the gene level in this dataset (Figure~\ref{sfig:snp_gene_manhattan}B), only 36 of all the genes retain a positive score after that transformation. Yet, this small solution does not provide much insight into the susceptibility mechanisms to cancer. Importantly, it ignores genes that are associated to cancer in this dataset like \emph{FGFR2}.

On the other end of the spectrum, dmGWAS, HotNet2, and SigMod produced large solutions. dmGWAS' subnetwork is the least associated subnetwork on average. This is due to the greedy framework it uses, which has a bias for larger solutions \citep{nikolayeva_network_2018}. It considered all nodes at distance 2 of the examined subnetwork, and accepted a weakly associated genes if it was linked to another, strongly associated one. This is exacerbated when the results of successive greedy searches are aggregated, leading to a large, tightly connected cluster of unassociated genes (Figure~\ref{fig:issues}A). This relatively low signal-to-noise ratio combined with the large solution requires additional analyses to draw conclusions, such as enrichment analyses. In the same line, HotNet2's subnetwork is even harder to interpret, being composed of 440 genes divided into 135 subnetworks. Lastly, SigMod misses some of the most strongly associated, breast cancer susceptibility genes in the dataset, like \emph{FGFR2} and \emph{TOX3}.

Another peculiarity of network methods is their relationship to degree centrality. On the one hand we observed that highly central genes often had no association to disease (Figure~\ref{fig:issues}B). On the other, network methods favor central genes, as they often connect high scoring nodes. This was specially the case of SigMod, which selected three highly central, unassociated genes: \emph{COPS5}, \emph{CUL3} and \emph{FN1}. As we showed in Section \ref{results:consensus}, and will show in \ref{results:stable-consensus}, there is evidence in the literature of the contribution of the first two to breast cancer susceptibility. With regards to \emph{FN1}, it encodes a fibronectin, a protein of the extracellular matrix involved in cell adhesion and migration. Overexpression of \emph{FN1} has been observed in breast cancer \citep{Ioachim2002}, and it is negatively correlated with poor prognosis in other cancer types \citep{Yi2016,Sponziello2016}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.9\linewidth]{../figures/sfigure_3.pdf}
  \caption{Pearson correlation between the different solution subnetworks. \textbf{(A)} Correlation between selected SNPs. \textbf{(B)} Correlation between selected genes. In general, the solutions display a very low overlap.}
  \label{sfig:pearson_methods}
  \end{figure}

By virtue of using a SNP subnetwork, SConES analyzes each SNP in their functional context. It therefore can select SNPs in genes without any associated interactor, as well as SNPs in non-coding regions or in non-interacting genes. In fact, due to linkage disequilibrium, SConES favors such genes, as selecting SNPs in an LD-block which overlaps with a gene favors selecting the rest of the gene. This might explain why SConES produces similar results on the GS and GM networks, heavily affected by linkage disequilibrium (Figure~\ref{sfig:pearson_methods}). On the other hand, SConES penalizes selecting SNPs and not their neighbors. This makes it conservative regarding SNPs with many interactions, like those mapped to hub genes in the PPIN. For this reason, SConES GI did not select any protein coding gene, despite selecting similar regions as SConES GS (Figure~\ref{fig:issues}C). In fact SConES GS and SConES GM select regions related to breast cancer, like 3p24 (\emph{SLC4A7}/\emph{NEK10} \citep{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \citep{quigley_5p12_2014}), 10q26 (\emph{FGFR2}), and 16q12 (\emph{TOX3}). On top of those SConES GS selects region 8q24 (\emph{POU5F1B} \citep{breyer_expressed_2014}). We hypothesize that the lack of results on the PPIN network of SConES GI and LEAN are due to the same cause: the absence of joint association of a gene and a majority of its neighbors. Although in the case of SConES other hyperparameters could lead to a more informative solution (e.g. a lower \(\lambda\) in Equation~\ref{eq:scones}), it is unclear what the best strategy to find them is. In addition, due to the design of the iCOGS array, the genome of GENESIS participants has not been unbiasedly surveyed: some regions are fine-mapped --- which might distort gene structure in GM and GI networks --- while others are under studied --- hindering the accuracy with which the GS network captures the genome structure.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{../figures/figure_5.pdf}
  \caption{Stable consensus subnetwork on GENESIS (Section~\ref{results:stable-consensus}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the most central genes (\emph{CUL3}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in Figure \ref{sfig:stable-consensus-names}.}
  \label{fig:stable-consensus}
\end{figure}
